---
title: "Analysis of Genetic Anomalies found on SINASC database"
author: "BioHub Solutions"
date: 03/09/2024
date-format: DD/MM/YYYY
format: html
editor: visual
code-fold: true
---

## Loading libraries

The main libraries are [read.dbc](https://github.com/danicat/read.dbc), [microdatasus](https://github.com/rfsaldanha/microdatasus) and [cid10](https://github.com/msrodrigues/cid10). Both are hosted on GitHub. Check the links to see how to install them. The other packages are for data manipulation, analysis and visualization.

```{r, warning=FALSE, message=FALSE}

#ETL
library(read.dbc)
library(microdatasus)
library(DBI)
library(RPostgres)

#data manipulation
library(tidyverse)
library(janitor)
library(cid10)

#data viz
library(ggplot2)
library(patchwork)
library(DT)
library(htmltools)
```

## Downloading data from dataSUS

What is dataSUS?....

```{r, warning=FALSE, message=FALSE}

#extract SINASC dataset
sinasc = fetch_datasus(year_start = 2014,
                       year_end = 2023,
                       month_start = 1,
                       month_end = 12,
                       uf = "PE",
                       information_system = "SINASC")

```

> NOTE: A characteristic of SINASC is that the dataset is release yearly and, as now, the data from 2023 is preliminary.

## Data manipulation

A few data cleaning, and transformation will be necessary to make the dataset more organized and accurate.

```{r}

#use the function process_sinac and clean names with janitor
sinasc = process_sinasc(sinasc) |> 
  clean_names()

#remove deprecated collumns remaining from the function process_sinasc

sinasc |> 
  mutate(locnasc = locnasc_as_factor_data_locnasc,
         idademae = idademae_as_factor_data_idademae,
         estcivmae = estcivmae_as_factor_data_estcivmae,
         escmae = escmae_as_factor_data_escmae, 
         gestacao = gestacao_as_factor_data_gestacao,
         parto = parto_as_factor_data_parto) |> 
  select(-c(contains("_as_factor_"))) |> 
  mutate(across(contador:codmunnasc, ~ as.numeric(.x))) |>
  mutate(locnasc = as.factor(locnasc)) |> 
  mutate(idademae = as.numeric(idademae)) |> 
  mutate(across(estcivmae:escmae, ~ as.factor(.x))) |>
  mutate(across(qtdfilvivo:codmunres, ~ as.numeric(.x))) |>
  mutate(across(gestacao:consultas, ~ as.factor(.x))) |>
  mutate(dtnasc = as.Date(dtnasc, tryFormats = "%Y-%m-%d")) |>
  mutate(sexo = as.factor(sexo)) |> 
  mutate(across(apgar1:apgar5, ~ as.numeric(.x))) |> 
  mutate(peso = as.numeric(peso)) |> 
  mutate(idanomal = as.factor(idanomal)) |> 
  mutate(dtcadastro = as.Date(dtcadastro, tryFormats = "%Y-%m-%d")) |>
  mutate(dtrecebim = as.Date(dtrecebim, tryFormats = "%Y-%m-%d")) |>
  mutate(across(naturalmae:codufnatu, ~ as.factor(.x))) |>
  mutate(across(escmae2010:seriescmae, ~ as.numeric(.x))) |> 
  mutate(dtnascmae = as.Date(dtnascmae, tryFormats = "%Y-%m-%d")) |>
  mutate(racacormae = as.factor(racacormae)) |>   
  mutate(across(qtdgestant:idadepai, ~ as.numeric(.x))) |>  
  mutate(dtultmenst = as.Date(dtultmenst, tryFormats = "%Y-%m-%d")) |>
  mutate(semagestac = as.numeric(semagestac)) |> 
  mutate(tpmetestim = case_when(tpmetestim == 1 ~ "exame físico",
                                tpmetestim == 2 ~ "outro método",
                                tpmetestim == 9 ~ "ignorado")) |> 
  mutate(tpmetestim = as.factor(tpmetestim)) |> 
  mutate(across(consprenat:mesprenat, ~ as.numeric(.x))) |>  
  glimpse()
```
